---
title: "NOS Table"
author: "Ronil V. Chandra"
date: "16/06/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Load packages

```{r load packages, echo = FALSE, include = FALSE}

library(tidyverse)
library(kableExtra)
```

### Load required data

A single excel data file is loaded to carry out all data table creation, with all steps documented below to ensure reproducibility of research.

```{r load data, echo = FALSE, include = FALSE}

nosdata <- read_csv("data-raw/NOS_LAS.csv")

nosdata <- nosdata %>%
  slice(-12)
```


### Calculate Quality Scores and convert to AHRQ thresholds

Quality assessment was performed using the Newastle Ottowa Scale as recommended by Cochrane; this is also the most commonly utilised for observational studies. 

This can be converted to the The Agency for Healthcare Research and Quality within the United States Department of Health and Human Services (AHRQ) standards using the following thresholds. 

Good quality: 
3 or 4 stars in selection domain AND 1 or 2 stars in comparability domain AND 2 or 3 stars in outcome/exposure domain

Fair quality: 
2 stars in selection domain AND 1 or 2 stars in comparability domain AND 2 or 3 stars in outcome/exposure domain

Poor quality: 
0 or 1 star in selection domain OR 0 stars in comparability domain OR 0 or 1 stars in outcome/exposure domain



Poor quality 

```{r calculate quality scoring as per AHRQ thresholds - poor quality}

nos.poor <- nosdata %>%
  filter(nos_select == 0 | nos_select == 1 | nos_compare == 0 | nos_outcome == 0 | nos_outcome == 1) %>%
  mutate(ahrq = "Poor")

```

Fair quality 

```{r calculate quality scoring as per AHRQ thresholds - fair quality}

nos.fair <- nosdata %>%
  filter(nos_select == 2) %>%
  filter(nos_compare == 1 | nos_compare == 2) %>%
  filter(nos_outcome == 2 | nos_outcome == 3) %>%
  mutate(ahrq = "Fair")


```


Good Quality 

```{r calculate quality scoring as per AHRQ thresholds - good quality}

nos.good <- nosdata %>%
  filter(nos_select == 3 | nos_select == 4 ) %>%
  filter(nos_compare == 1 | nos_compare == 2) %>%
  filter(nos_outcome == 2 | nos_outcome == 3) %>%
  mutate(ahrq = "Good") 

```

Join tables

```{r join tibbles, and arrange according to AHRQ}

nos.temp <- full_join(nos.poor, nos.fair)

nos.final <- full_join(nos.temp,nos.good) %>%
  unite(auth_year, c(auth, pub), sep = " ", remove = FALSE) %>%
  select(auth_year, sel_1, sel_2, sel_3, sel_4, nos_compare, out_1, out_2, out_3, nos_total, ahrq) %>%
  arrange(match(ahrq, c("Good", "Fair", "Poor")), desc(nos_total))

```


```{r recode variables to star system for NOS}

nos.final$sel_1 <- as.character(nos.final$sel_1)
nos.final$sel_1 <- recode(nos.final$sel_1, "1" = "*", "0" = " ")

nos.final$sel_2 <- as.character(nos.final$sel_2)
nos.final$sel_2 <- recode(nos.final$sel_2, "1" = "*", "0" = " ")

nos.final$sel_3 <- as.character(nos.final$sel_3)
nos.final$sel_3 <- recode(nos.final$sel_3, "1" = "*", "0" = " ")

nos.final$sel_4 <- as.character(nos.final$sel_4)
nos.final$sel_4 <- recode(nos.final$sel_4, "1" = "*", "0" = " ")

nos.final$nos_compare <- as.character(nos.final$nos_compare)
nos.final$nos_compare <- recode(nos.final$nos_compare, "1" = "*", "2" = "* *", "0" = " ")

nos.final$out_1 <- as.character(nos.final$out_1)
nos.final$out_1 <- recode(nos.final$out_1, "1" = "*", "0" = " ")

nos.final$out_2 <- as.character(nos.final$out_2)
nos.final$out_2 <- recode(nos.final$out_2, "1" = "*", "0" = " ")

nos.final$out_3 <- as.character(nos.final$out_3)
nos.final$out_3 <- recode(nos.final$out_3, "1" = "*", "0" = " ")

```

### Quality Assessment Table


```{r, Quality Table, results = 'asis'}

nos.final %>%
  
  knitr::kable(
    caption = "Risk of bias assessment for non-randomized studies", 
    col.names = c("Study", "Representativeness of the exposed cohort", "Selection of the non-exposed cohort", "Ascertainment of prior subarachnoid haemorrhage", "Demonstration that outcome of interest was not present at start of study", "Comparability of cohorts on the basis of the design or analysis", "Assessment of outcome", "Was follow-up long enough for outcomes to occur?", "Adequacy of follow up of cohorts", "NOS Quality Score", "AHRQ Standards"),
    align = c("l", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c")
    ) %>%
  kableExtra::kable_styling(bootstrap = "striped", 
                            full_width = FALSE,
                            font_size = 10,
                            ) %>%
  pack_rows("AHRQ Good Standard", 1, 10) %>%
  pack_rows("AHRQ Fair Standard", 11, 12) %>%
  pack_rows("AHRQ Poor Standard", 13, 30) %>%
  
  add_header_above(c(" " = 1, "Selection" = 4, "Comparability" = 1, "Outcome" = 3, "Total Scores" = 2 )) %>%
  footnote(
    general_title = "Footnote:",
    general = "Thresholds for converting the Newcastle-Ottawa scales to AHRQ standards (good, fair, and poor): 
    Good quality: 3 or 4 stars in selection domain AND 1 or 2 stars in comparability domain AND 2 or 3 stars in outcome/exposure domain
    Fair quality: 2 stars in selection domain AND 1 or 2 stars in comparability domain AND 2 or 3 stars in outcome/exposure domain
    Poor quality: 0 or 1 star in selection domain OR 0 stars in comparability domain OR 0 or 1 stars in outcome/exposure domain",
    threeparttable=TRUE,
    )


```







